Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2020-05-05 code/main.tex
starting reply to reviewer

\begin{abstract}

\end{abstract}
%
\begin{keywords}
keyword1, keyword2
\end{keywords}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

On the particular case of the difusion of signals of intelligence in
the Galaxy, the discrete events method can be performed taking into
account a small number of variables.  The temporal structure is
defined by two distribution parameters that represent the mean
time interval that an intelligent civilization can emit and receive
signals, and the mean time interval between the emergence of an
intelligent comunicating civilization and the next.
The spatial structure of the simulation is given by the size and shape
of the Galactic Habitable Zone and the maximum distance a signal can
travel to be detected.

Also, some hypothesis must be made in order to complete the
simulation, namely:

\begin{itemize}
   \item The capacity to emit signals and to receive signals ocurr at
      the same time.
   \item All CETIs use the same signal power, so that there is a
      maximum distance out to which it can be detected.
   \item The distribution of the appeareance of new CETIs
      is a stationary Poisson distribution.
   \item The distribution of duration of CETIs
      is a stationary exponential distribution.
   \item There is no a temporal window 
   \item The probability for the rise of a CETI is homogeneous
      over the GHZ.
   \item The message travels at light speed.

\end{itemize}

@@

\begin{itemize}
   \item Size of the Galactic Habitable Zone.  Radial symetry is
      assumed, so that the GHZ is determined by two radii, the inner
      radius (GHZ_inner) and the outer radius (GHZ_outer).  They are
      measured in light years, with the aim to maintain y single and
      comprehensive unit for both space and time coordinates.
   \item Mean lifetime of a CETI
   \item Mean waiting time for the appeareance of another CETI
   \item Maximum distance a message can be detected
\end{itemize}

@@

%#
%# CONTENTS OF THE <CETIS> PICKLE:
%# . . . . . . . . . . . . . . . . . . . . . . . .
%# 0. ID of emitting CETI
%# 1. ID of emitting [receiving] CETI
%# 2. x position in the galaxy
%# 3. y position in the galaxy
%# 4. time of Awakening [Contact]
%# 5. time of Doomsday  [Blackout]
%# . . . . . . . . . . . . . . . . . . . . . . . .
%#
%# This structure is always for the A-D cycle of each emitting CETI,
%# and repeats for each contact (C-B).
%#
%# The time span of a CETI is t_D - t_A for CETIs[k][0]
%# The time span of a contact is t_B - t_C for CETIs[k][i], with i>0
%#
%#
%# age: time elapsed from A to a given time
%# ceti_e: ceti emitting signals (emiter)
%# ceti_r: ceti receiving signals (receiver)
%# ceti_c: ceti that listen at least another ceti (citizen)
%# ceti_h: ceti that is lestened by at least another ceti
%#
%
%# for Check purposes
%#-------------------
%# duration of a civilization (exponential distribution by costruction)
%# time from the appeareance of a CETI to the next
%#
%# 
%# derived quantities
%#-------------------
%# [time]
%# time span of a ceti listening another
%# time span of a ceti being listened by another
%# duration of two-way communication channels
%# waiting time until the first contact
%# fraction of awaken time a ceti is listening at least another ceti
%# age of contacted cetis_e at first C
%# distribution of time to wait until next C
%# fraction of cetis where the first contact is given at the A
%
%# [numbers]
%# distribution of the number of contacts for each ceti
%# distribution of the maximum number of contacts for each ceti
%# distribution of the number of contacts as a function of age
%# number of contacts as a function of time in the galaxy
%# rate of cetis that succeed in contact 
%# fraction of contacts that admit a response
%#
%# [distances]
%# distribution of distances between contacted cetis
%#
%# [correlations] relation between {RB}
%# RB distance to ceti and time of double communication
%# RB distance to ceti and age of contacted ceti
%# RB the age of a ceti and the maximum number of contacted cetis before D
%# RB lifespan of a ceti and max number of contacts
%#
%# distribution in the galaxy of cetis that reach contact
%
%# ... everything as a function of parameters!

Output
\begin{itemize}
   \item La salida de la simulación está guardad en una variable llamada CETI.
   \item La misma es una lista de objetos que son también listas.
   \item El elemento cero de cada lista (i.e., cada elemento de CETI) contiene:
   \item coordenada X
   \item coordenada Y
   \item tiempo de inicio de la capacidad de comunicación
   \item tiempo de final de la capacidad de comunicación
   \item y los demás elementos contienen:
   \item coordenada X de la CETI
   \item coordenada Y de la CETI
   \item tiempo del primer contacto
   \item tiempo del último contacto
\end{itemize}

@@

\section{Results}
%{{{

Resultados cuantitativos de la simulación
\begin{itemize}
   \item distribucion del numero de contactos
   \item distribucion de la duracion de los contactos
   \item distribucion del tiempo que hay que esperar para el proximo contacto
\end{itemize}

@@

Posibles implicaciones
\begin{itemize}
   \item Análisis del método de comunicación: isotrópico, colimado, accidental
   \item Naturaleza del transporte del mensaje: radiación electromagnética u otros
   \item Efectos de fenómenos de alineación (tránsitos)
   \item Marcadores energéticos
   \item Marcadores espirituales
   \item Uso de estrellas como amplificadores o fuentes
\end{itemize}

>>>>

\begin{abstract}
%
The abundance of intelligent civilizations in the galaxy is a
longstanding question, which is often conceptualized as the problem
of the lack of received communication of the Fermi paradox.
%
Most efforts on the estimation of a number of intelligent
civilizations are centered on the Drake equation, although 
its factors are affected by large uncertainties, and it lacks a
temporal nature.
%
Here we present numerical simulations of stochastic processes of
emergence of civilizations with communication capabilities, and
the causal contacts among them.
%
We analyze the rate of causal contacts as a function of the mean
number of civilizations and the mean lifetime span distribution.
%
Our results indicate that, given the large distances involved, in
   the odds to obtain a contact in a few years of monitoring, assuming a
   perfect detection rate, are very rare.
\end{abstract}
%
\begin{keywords}
keyword1, keyword2
\end{keywords}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

For citation purposes \citep{Lineweaver2004}
For citation purposes \citep{Forgan2011}
For citation purposes \citep{Rahvar2016}
For citation purposes \citep{Glade2011}
For citation purposes \citep{Dayal2016}
For citation purposes \citep{Gobat2016}
For citation purposes \citep{Prantzos2013}
For citation purposes \citep{Horvat2007}
For citation purposes \citep{Starling2013}
For citation purposes \citep{Barlow2012}
For citation purposes \citep{Tarter2009}
For citation purposes \citep{Forgan2017}
For citation purposes \citep{Forgan2013}
For citation purposes \citep{Wright2015}
For citation purposes \citep{Loeb2006}
For citation purposes \citep{Solomonides2016}
For citation purposes \citep{Fogg1987}
For citation purposes \citep{Anchordoqui2017}
For citation purposes \citep{Balb2018}
For citation purposes \citep{Forgan2010}
For citation purposes \citep{Forgan2008}
For citation purposes \citep{Forgan2016}
For citation purposes \citep{Haqq-Misra2017}
For citation purposes \citep{Forgan2015}
For citation purposes \citep{Gleiser2010}
For citation purposes \citep{cirkovic_temporal_2004}

For citation purposes \citep{cirkovic_temporal_2004}

A common approach to discuss the possible number of extraterrestrial
intelligent civilizations (ETI) has been the use of the Drake equation
\citep{}.

Uncertainties in the factors

Stochastic Drake equation and probabilistic approaches

Time:
The Impact of the Temporal Distribution of Communicating Civilizations on Their Detectability
Temporal dispersion of the emergence of intelligence: an inter-arrival time analysis
The spatiotemporal aspects of SETI
\citep{Fogg1987, Forgan2011, Balb2018}

Effort on considering the stochastic nature of the Drake equation
\citep{Glade2011}.

Simulation approach
New numerical determination of habitability in the Galaxy: the SETI connection
\citep{Forgan2008, Forgan2010}

@@

Some complex stochastic processes can be efficiently modeled with the
discrete--event (DE) simulation approach.
%
Simulations are suitable tools to analyze systems that evolve with
time and involve randomness.
%
In general, simulations require less assumptions and simplifications
than theoretical approaches, and can be applied to systems where 
a theoretical model can not be found.
%
A system described with the DE paradigm is characterized by a set of
actors and events, where actors interact causally through a series of
events on a time line and process these events in chronological order
\citep{ptolemaeus_book, chung_book, simulation_book}.
%
On the particular case of the difusion of signals of intelligence in
the Galaxy, the discrete events method can be performed taking into
account a small number of variables.
%
The temporal structure is defined by two distribution parameters that
represent the mean time interval that an intelligent civilization can
emit and receive signals, and the mean time interval between the
emergence of an intelligent comunicating civilization and the next one.
%
The spatial structure of the simulation is given by the size and shape
of the Galactic Habitable Zone and the maximum distance a signal can
travel to be detected.
%
The parameters for the temporal distributions also determine the
spatial properties.
%
For example, the density of active cetis in the Galaxy depend on these
two parameters.
%
Also, some hypothesis must be made in order to complete the
simulation, namely:

\begin{itemize}
   \item The capacity to emit signals and to receive signals ocurr at
      the same time.
   \item All CETIs use the same signal power, so that there is a
      maximum distance out to which it can be detected.
   \item The distribution of the appeareance of new CETIs
      is a stationary Poisson distribution.
   \item The distribution of duration of CETIs
      is a stationary exponential distribution.
   \item There is no a temporal window 
   \item The probability for the rise of a CETI is homogeneous
      over the GHZ.
   \item The message travels at light speed.
\end{itemize}

The exponential distribution of lifespan and waiting times is
justified by considering that the process of appeareance of life in
the galaxy is homogeneous and stationary.
%
This means that there is no preferred location within the GHZ for the
spontaneos appeareance of life, and that the emergence of a
civilization is independent of the existence of previous civilizations 
in the galaxy.
%
This is equivalent to proposing a Poisson process for the emergence of
CETIs, since there is a close relation between the number of events in
time or space and the waiting time or separation, respectively.
%
That is, these are two alternative approaches to describing the same
process, a Poisson distribution for the number of events implies an
exponential distribution for their sepations, and viceversa.
%
It should be emphasized that the exponential laws used in this work
are assumed, although instead of analyzing results from a particular
parameter chosen ad--hoc, we explore the hypothesis space and analyze
the impact of the values of these parameters on the results.
%

\subsection{Power laws vs. exponential laws}

The power law and exponential statistical distributions are among the
most common patterns found in natural phenomena.
%
For example, the distribution of the frecuency of words in many
languajes is known to follow a Zipf law (which is a power law).
%
Zipf law also describes population ranks of cities in various
countries, corporation sizes, income rankings, ranks of number of
people watching the same TV channel (Zipf's law entry on Wikipedia).
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and
floods; the sizes of meteorites or the losses caused by
business interruptions from accidents, are also well described by power
laws (Sornette 2006).
%
\itshape
Other studies have
documented power laws in stock market fluctuations,
sizes of computer files and word frequency in languages
(Mitzenmacher, 2004; Newman, 2005; Simkin \& Roy-
chowdhury, 2006).
In biology, power laws have been particularly impor-
tant in analysing connectivity patterns in metabolic
networks (Barabasi \& Albert, 1999; Ravasz et al., 2002)
and in the number of species observed per unit area in
ecology (Garcia Martin \& Goldenfeld, 2006).
(VER Frank, the common patterns of nature).
\slshape

Also, exponential laws arise when...

\subsection{Exploration of the parameter space}

Zipf law.

The adopted values for the GHZ are XX and XX \citep{Lineweaver2004}.

\begin{table}
\begin{tabular}{ccc}
\hline
   parameter & \multicolumn{2}{c}{adopted value}\hfill \\
\hline
   inner GHZ radius & \multicolumn{2}{c}{}\\
   outer GHZ radius & \multicolumn{2}{c}{}\\
\hline
   parameter & min & max \\
\hline
   D$_{max}$ & & \\
   $\tau_{a}$ & & \\
   $\tau_{N}$ & & \\
\hline
\end{tabular}
\caption{Adopted values for fixed and variable parameters in the
   simulation}
\label{T_parameters}
\end{table}

@@

\begin{itemize}
   \item Size of the Galactic Habitable Zone.  Radial symetry is
      assumed, so that the GHZ is determined by two radii, the inner
      radius (GHZ\_inner) and the outer radius (GHZ\_outer).  They are
      measured in light years, with the aim to maintain y single and
      comprehensive unit for both space and time coordinates.
   \item Mean lifetime of a CETI
   \item Mean waiting time for the appeareance of another CETI
   \item Maximum distance a message can be detected
\end{itemize}

@@

Direct quantities are:

\begin{itemize}
   \item ID of emitting CETI
   \item ID of emitting (receiving) CETI 
   \item x position in the galaxy
   \item y position in the galaxy
   \item time of Awakening (Contact)
   \item time of Doomsday  (Blackout)
\end{itemize}

This structure is always for the A-D cycle of each emitting CETI,
and repeats for each contact (C-B).
%
We define the following quantities: age is the time elapsed from A to a
given time, ceti\_e is the CETI emitting signals (emiter), ceti\_r is
the CETI receiving signals (receiver), ceti\_c is the ceti that listen
at least another ceti (citizen) and ceti\_h is the CETI ceti that is
lestened by at least another CETI.

Then, several derived quantities can be obtained:

\begin{itemize}
   \item time span of a ceti listening another
   \item time span of a ceti being listened by another
   \item duration of two-way communication channels
   \item waiting time until the first contact
   \item waiting time until the next contact
   \item fraction of awaken time a ceti is listening at least another ceti 
   \item age of contacted cetis\_e at first C
   \item distribution of time to wait until next C
   \item fraction of cetis where the first contact is given at the A       
   \item distribution of the number of contacts for each ceti         
   \item distribution of the maximum number of contacts for each ceti 
   \item distribution of the number of contacts as a function of age  
   \item number of contacts as a function of time in the galaxy       
   \item rate of cetis that succeed in contact                        
   \item fraction of contacts that admit a response                   
   \item distribution of distances between contacted cetis 
   \item relation between distance to ceti and time of double communication 
   \item relation between distance to ceti and age of contacted ceti
   \item relation between the age of a ceti and the maximum number of contacted cetis before D
   \item relation between lifespan of a ceti and max number of contacts
   \item distribution in the galaxy of cetis that reach contact
\end{itemize}

All these quantities can be analyzed as a function of the simulation
parameters.

@@

\subsection{Duration of contacts}
%{{{

\subsection{Waiting time for a first contact}
%{{{

In this section we analyze the distribution of the waiting times for
a first contact.
%
Such distribution can be considered to compute the probability of
reaching a first contact within the first XX years.

\subsection{Reciprocal communication}
%{{{

\subsection{Location within the Galaxy}
%{{{

@@

\section{Discusion}
%{{{

Possible implications of the results:

\begin{itemize}
   \item Analysis of the communication method: isotropic, colimated,
      serendipitous
   \item Nature of the message carrier: electromagnetic or another
   \item Effects of alignments
   \item energetic markers
   \item Spiritual markers
   \item Use of stars as sources or amplifiers
\end{itemize}

### 2020-05-05 submit_v02/manuscript.tex
starting reply to reviewer

\jname{Astrobiology}%

@@

\input{captions}

@@

\begin{figure}[!t]
   \centering
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   \usecaption{F_scheme} \label{F_scheme}
\end{figure}

@@

\begin{figure*}
   \centering
   \includegraphics[width=\textwidth]{F_messages.pdf}
   \usecaption{F_messages} \label{F_messages}
\end{figure*}

@@

\begin{figure} \centering
   \includegraphics[width=0.5\textwidth]{F_number_of_contacts.pdf}
   \usecaption{F_number_of_contacts} \label{F_number_of_contacts}
\end{figure}

@@

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_never_contact.pdf}
   \usecaption{F_never_contact}
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_C_at_A.pdf}
   \usecaption{F_C_at_A}
   \label{F_C_at_A}
\end{figure*}

@@

\begin{figure}
   \centering
   \includegraphics[width=0.5\textwidth]{F_waiting_for_1C.pdf}
   \usecaption{F_waiting_for_1C}
   \label{F_waiting_for_1C}
\end{figure}

>>>>

\jname{International Journal of Astrobiology}%

@@

@@

\begin{figure}[!t]
   \centering
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   \caption{
   %
Space--time diagrams showing a schematic representation of the
different stages in the development of causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (left vertex), following an "A" type event
("Awakening") in which the node acquires the communication capability.
%
The sphere of first contact (SFC) is a sphere centered on the emitter
that grows until its radius reaches the D$_{max}$ distance, in which
the power of a signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the
region $R$.
%
The surface of last contact (SLC) is another sphere that grows from a
"D" event ("Doomsday").
%
The region which is causally connected to the emitter is then limited
by these two spheres, and has the shape of a sphere or of a spherical
shell, depending on the time.
%
The temporal intervals for
the communication between two \cetis{} are represented in he
panel (b).
%
The receiver \ceti{} E$_1$ can listen signals from the emiter \ceti{} E$_2$,
from the "Contact" event (t=C$_{12}$) up to the "Blackout" event (t=B$_{12}$).
%
Similarly, 
the receiver \ceti{} E$_2$ can listen signals from the emiter \ceti{} E$_1$,
from t=C$_{21}$ up to t=B$_{21}$. 
%
   } \label{F_scheme}
\end{figure}

@@

\begin{figure*}
   \centering
   \includegraphics[width=\textwidth]{F_messages.pdf}
   \caption{
   %   
Schematic representation on space--time diagrams of the possible cases in which a
\ceti{} E (Emiter, solid line) can be in causal contact with a \ceti{} R
(Receiver, dashed line).
%
Causal contact can be produced either if the emiter \ceti{} appears before
(left) or after (right) the receiver \ceti{}.
%
The causal contact in which the receiver can have access to the signal 
from the emiter is produced between the C event (Contact, open circles) and the B
event (Blackout, filled circles).
%
The duration of the causal contact in one direction depends on
several factors, mainly the time lag between the awakening events
(horizontal separation in the space--time diagrams)
and the distance between the \cetis (vertical separation in the
space--time diagrams).
%   
The different configurations must be taken into account in order to 
obtain the list of the upcomming events.
%
The time intervals for a two--way communication to be possible are
indicated in double lines. 
%
   } \label{F_messages}
\end{figure*}

@@

\begin{figure} \centering
   \includegraphics[width=0.5\textwidth]{F_number_of_contacts.pdf}
   \caption{
   %
Empirical cumulative distributions of the number contacts
for \cetis in four different samples (M5, M6, M7, M8) with two
values for the range of the signal reach: D$_{max}$=40000 ly (blue),  and 
D$_{max}$=80000 ly (orange).
%
Except for the models corresponding to a densely populated galaxy with
long--living \cetis, the \cetis which receive more than 10 contacts
are very unlikely.
%
See Table \ref{T_selected_models} for model descriptions.
%
   } \label{F_number_of_contacts}
\end{figure}

@@

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_never_contact.pdf}
   \caption{
   %
Rate of MPLs that never make contact (listening) as a
function of $\tau_a$ and $\tau_s$, for 
$D_{max}$=10000 (left panel),
$D_{max}$=40000 (middle panel), and
$D_{max}$=80000 (right panel)
%
   }
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_C_at_A.pdf}
   \caption{
   %
Rate of MPLs that listen at the moment of the
awakening, as a funcion of $\tau_a$ and $\tau_s$, for
$D_{max}$=10000 (left panel),
$D_{max}$=40000 (middle panel), and
$D_{max}$=80000 (right panel)
%
   }
   \label{F_C_at_A}
\end{figure*}

@@

\begin{figure}
   \centering
   \includegraphics[width=0.5\textwidth]{F_waiting_for_1C.pdf}
   \caption{
   %
Histograms of the mean waiting times for the first contact, for
several models.
%
Left panel shows the histograms for several values of $\tau_a$, and
$\tau_s$ in the range 17000-24000 yr.
%
Right panel shows the histograms for several values of $\tau_s$, and
$\tau_a$ in the range 68000-100000 yr.
%
The shape of the histograms does not change when $\tau_a$ is
modified, but significantly changes when $\tau_s$ is modified. 
%
   }
   \label{F_waiting_for_1C}
\end{figure}
